---
title: "Variable Exploration"
author: "Jennifer Collister"
date: "`r Sys.Date()`"
output: 
  word_document:
      fig_width: 8
      fig_height: 5
      reference_docx: "K:/TEU/TEU_Guides/TEU_DocStyle_Rmd_2020.dotx"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(include=TRUE, echo=FALSE, warning = FALSE, message = FALSE) 

library(kableExtra)
library(knitr)
# library(yaml)
library(here)
library(ggplot2)
library(tidyr)
# library(readr)
library(dplyr)
# library(glue)
# library(BCRA)

# Pretty tables
# library(arsenal)
# library(pander)

# Survival 
# library(survival)
# library(survminer)
# library(rms)


# Set the root directory to the project directory (otherwise knitr works in the directory of the Rmd file by default)
knitr::opts_knit$set(root.dir = here::here())

# Load the project config file for filepaths etc
# config = yaml.load_file(here::here("./config.yml"))
# source(here::here(file.path(config$functions)))

#-------------------------------------------------------------------------------------------------------------------

# Output Formatting

# Specify the markdown format for knitr tables, otherwise they're not compatible with kableExtra
options(knitr.kable.NA='', knitr.table.format = "markdown")
options(kableExtra.auto_format = FALSE)

#-------------------------------------------------------------------------------------------------------
# Setup

figure <- 1
table <- 1
suppfig <- 1
footnote_no <- 1

```


```{r derive-data, include=FALSE, cache=TRUE}

# source(file.path(config$scripts$cleaning, "dataset_generator.R"))

exclusions <- function(data) {
  
  excl$initial <<- nrow(data)
  
  # Restrict to females (by self-report and genetic)
  data <- data[data$BaC_Sex == "Female",]
  data <- data[data$GeP_Sex == "Female" & !is.na(data$GeP_Sex),]
  excl$sex <<- nrow(data)
  
  # Restrict to post-menopausal women
  data <- data[!is.na(data$FSF_MenopauseStatus) & data$FSF_MenopauseStatus == "Yes",]
  excl$meno <<- nrow(data)
  
  # Restrict to White British (genetically)
  data <- data[!is.na(data$GeP_ethnic),]
  excl$whitebrit <<- nrow(data)
  
  # Exclude those outside the 40-70 age range
  data <- data[!is.na(data$TEU_BaC_AgeAtRec) & data$TEU_BaC_AgeAtRec >= 40 & data$TEU_BaC_AgeAtRec < 70,]
  excl$agerange <<- nrow(data)
  # Restrict population to those for whom we have PRS scores available
  data <- data[!is.na(data$PRS_bc_std),]
  excl$prs <<- nrow(data)
  # Exclude women with 20+ sisters as the TC model can't handle it
  data <- data[is.na(data$TC_NSisters) | data$TC_NSisters < 20,]
  # Exclude those who didn't answer HRT use
  data <- data[data$TC_HRT_raw != "Unanswered",]
  # Or who reported impossible HRT dates
  data <- data[is.na(data$FSF_HRTAgeStop) | data$FSF_HRTAgeStop <= ceiling(data$TEU_BaC_AgeAtRec),]
  excl$hrt <<- nrow(data)
  
  # Exclude prior breast cancer
  data <- data[is.na(data$BrCaDx_Prevalent),]
  data <- data[is.na(data$VI_BrCa),]
  excl$brca <<- nrow(data)
  # Prior mastectomy
  data <- data[data$TEU_HES_Mast_prev == "No",]
  excl$mast <<- nrow(data)
  # Prior carcinoma in situ
  data <- data[is.na(data$BrCaInSitu_Prevalent),]
  excl$cainsitu <<- nrow(data)

  
  return(data)
}

excl <- list(initial=0)

# data <- derive_variables(
#   database = config$data$database,
#   field_definitions = TEU_SPECS$TyrerCuzick,
#   exclusions = function(data){return(data[data$BaC_Sex == "Female" &
#                                             data$GeP_Sex == "Female" & !is.na(data$GeP_Sex),])}
#   )
# saveRDS(data, "K:/TEU/PRSforBrCa/Stats_Outputs/agestdinc.rds")
data <- readRDS("K:/TEU/PRSforBrCa/Stats_Outputs/agestdinc.rds")

wholepop <- data
analysis_pop <- exclusions(data)

```


# Age-specific and directly standardised breast cancer incidence rates

> We calculated age specific and directly standardised colorectal cancer incidence rates in UK Biobank overall and for the Integrated Modelling Cohort used to derive integrated QCancer-10+PRS risk models, and compared these with Office for National Statistics 2013 cancer registry data for England (chosen as the approximate mid-point of available UK Biobank follow-up).[15](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/datasets/cancerregistrationstatisticscancerregistrationstatisticsengland) Age specific rates were calculated in five year age bands between 40 and 80 years as the number of first incident colorectal cancers over the number of person years at risk. Age standardised incidence rates were calculated using the 2013 European Standard Population aged 40-80 years.[16](https://webarchive.nationalarchives.gov.uk/ukgwa/20160105231503/http://www.ons.gov.uk/ons/guide-method/user-guidance/health-and-life-events/index.html) Rates are presented per 100 000 person years at risk (supplementary methods). - Briggs2022

>  Cancer Incidence Calculation 
Whole UKB cohort CRC incidence rates were calculated based on linked registry cases, without removal of prevalent cases, to reflect registration as would occur in national data. In addition ASIRs were calculated in the Integrated Modelling Cohort, in which prevalent cases were removed and cases identified through cancer and death registry, and linked hospital inpatient data; follow-up duration was as defined in the main methods. This analysis used R packages ‘survival’ and ‘epitools’. - [Briggs et al., 2022, p. 4]

The tables from Briggs2022 citations 15 and 16 are saved at K:\TEU\PRSforBrCa\Resources\referencetablearv2013c44_tcm77-410534.xls and K:\TEU\PRSforBrCa\Resources\agestandardisedmortalityratecalculationtemplateusingthe2013esp_tcm77-359944.xls respectively.

## UK Biobank overall

Prevalent and incident cases are included.

```{r}

# UKB overall (ie including prevalent cases)

# Our TEU_BrCa_status and TEU_BrCa_age are generated for prevalent cases
# So supplement with BrCaAge_Any (age at brca dx from Cancer Registry) to include the incident cases
# Note TEU_BrCa_age is the age at which follow-up ended, ie either breast cancer dx or censoring (death by other causes, lost to followup, admin censoring)

data <- wholepop %>% transmute(
  TEU_BrCa_status = as.numeric(!is.na(BrCaAge_Any) | TEU_BrCa_status), # Include prevalent cases
  TEU_BrCa_age = coalesce(BrCaAge_Any, TEU_BrCa_age), # Take CaR recorded age preferentially if it exists
  TEU_BaC_AgeAtRec = TEU_BaC_AgeAtRec
) %>%
  filter(TEU_BrCa_age >= 40) # Only interested in cancers from age 40

groups <- c()
brca_n <- c()
atrisk_n <- c()

for(age in c(55, 60, 65, 70, 75, 80)){
  
  # Define the 5-yr interval of interest
  start <- age
  end <- age + 5
  
  # Prep labels
  group <- paste0("[", start, ",", end, ")")
  groups <- c(groups, group)
  
  # Number of cases in this age group (had breast cancer and age at dx within this interval)
  brca <- data %>%
    filter(TEU_BrCa_status == 1 & TEU_BrCa_age >= start & TEU_BrCa_age < end)
  brca_n <- c(brca_n, nrow(brca))
  
  # Number of person years at risk in this age group
  atrisk <- data %>% 
    filter(TEU_BrCa_age >= start) %>% # If they weren't censored before the start of this interval
    mutate(start_int = start,
           end_int = pmin(TEU_BrCa_age, end),
           time_at_risk = end_int - start_int
           )
  
  atrisk_n <- c(atrisk_n, sum(atrisk$time_at_risk))
}

rates_wholepop <- data.frame(Age5 = groups,
                    brca = brca_n,
                    n = atrisk_n,
                    esp_2013 = c(6500, 6000, 5500, 5000, 4000, 2500)) %>%
  mutate(Age_specific_rates = brca/(n/100000),
         lci = 0.5 * qchisq(p=0.025, df=(2*brca))/(n/100000),
         uci = 0.5 * qchisq(p=0.975, df=(2*(brca+1)))/(n/100000),
         std_wt = esp_2013/sum(esp_2013),
         Age_standardised_rates = std_wt * Age_specific_rates)

sum(rates_wholepop$Age_standardised_rates)


kable(rates_wholepop, digits=2)

ASR <- sum(rates_wholepop$Age_standardised_rates)
LCI <- round(ASR - 1.96*(ASR/sqrt(sum(rates_wholepop$brca))), 2)
UCI <- round(ASR + 1.96*(ASR/sqrt(sum(rates_wholepop$brca))), 2)

```

The age-standardised breast cancer rate per 100,000 population (ages 55-85) is: `r round(ASR,2)` (95% CI: `r LCI`, `r UCI`).

This compares to the ONS 2013 directly age-standardised rates per 100,000 population of 299.5 for women.

## Analysis population

Population is post-menopausal White British women with no history of breast cancer, carcinoma in situ or risk reducing mastectomy.

Only incident cases are considered.

```{r}

# Analysis population

data <- analysis_pop %>% select(
  TEU_BrCa_status,
  TEU_BrCa_age,
  TEU_BaC_AgeAtRec
)

groups <- c()
brca_n <- c()
atrisk_n <- c()

for(age in c(55, 60, 65, 70, 75, 80)){
  
  start <- age
  end <- age + 5
  
  group <- paste0("[", start, ",", end, ")")
  groups <- c(groups, group)
  
  # Number of cases in this age group (had breast cancer and age at dx within this interval)
  brca <- data %>% 
    filter(TEU_BrCa_status == 1 & TEU_BrCa_age >= start & TEU_BrCa_age < end)
  brca_n <- c(brca_n, nrow(brca))
  
  # Number of person years at risk in this age group
  atrisk <- data %>% 
    filter(TEU_BaC_AgeAtRec < end & TEU_BrCa_age >= start) %>% # Risk only starts at baseline in analysis pop
    mutate(start = pmax(TEU_BaC_AgeAtRec, start),
           end = pmin(TEU_BrCa_age, end),
           time_at_risk = end - start
           )
  
  atrisk_n <- c(atrisk_n, sum(atrisk$time_at_risk))
}

rates_analysispop <- data.frame(Age5 = groups,
                    brca = brca_n,
                    n = atrisk_n,
                    esp_2013 = c(6500, 6000, 5500, 5000, 4000, 2500)) %>%
  mutate(Age_specific_rates = brca/(n/100000),
         lci = 0.5 * qchisq(p=0.025, df=(2*brca))/(n/100000),
         uci = 0.5 * qchisq(p=0.975, df=(2*(brca+1)))/(n/100000),
         std_wt = esp_2013/sum(esp_2013),
         Age_standardised_rates = std_wt * Age_specific_rates)

sum(rates_analysispop$Age_standardised_rates)


kable(rates_analysispop, digits=2)

ASR <- sum(rates_analysispop$Age_standardised_rates)
LCI <- round(ASR - 1.96*(ASR/sqrt(sum(rates_analysispop$brca))), 2)
UCI <- round(ASR + 1.96*(ASR/sqrt(sum(rates_analysispop$brca))), 2)

```

The age-standardised breast cancer rate per 100,000 population (ages 55-85) is: `r round(ASR,2)` (95% CI: `r LCI`, `r UCI`).

This compares to the ONS directly age-standardised rates per 100,000 population of 299.5 for women.

# Plot age-specific breast cancer rates, compare to ONS


```{r}

# Plot results from
# Analysis population: above chunk
# Whole UKB population: branch wholepop
# ONS 2013 data: K:\TEU\PRSforBrCa\Resources\referencetablearv2013c44_tcm77-410534.xls

rateplot <- cbind(rates_analysispop %>% transmute(Age5 = Age5, 
                                                  rate_analysis = Age_specific_rates,
                                                  lci_analysis = lci,
                                                  uci_analysis = uci),
                  rates_wholepop %>% transmute(rate_wholeUKB = Age_specific_rates,
                                               lci_wholeUKB = lci,
                                               uci_wholeUKB = uci),
               rate_ONS = c(276.0, 354.8, 415.9, 354.9, 400.6, 432.1)
               ) %>%
  mutate(lci_ONS = rate_ONS,
         uci_ONS = rate_ONS) %>%
  pivot_longer(cols=c(rate_analysis, rate_wholeUKB, rate_ONS,
                      lci_analysis, lci_wholeUKB, lci_ONS,
                      uci_analysis, uci_wholeUKB, uci_ONS),
               names_to = c(".value", "name"),
               names_sep = "_") %>%
  mutate(Cohort = factor(name, levels=c("ONS", "wholeUKB", "analysis"),
                       c("ONS 2013", "Whole UK Biobank", "Analysis population")))

ggplot(data=rateplot, aes(x=Age5, y=rate, group=Cohort, color=Cohort)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=lci, ymax=uci), width=.3, position=position_dodge(0.1)) +
  ggtitle("Age specific breast cancer rates in women in the UK Biobank,\noverall and within analysis cohort, compared to\nOffice for National Statistics 2013 Cancer Registry data") +
  ylab("Incidence rate per 100,000 person years at risk") +
  xlab("Age at cancer diagnosis, years") 

# ggsave(file.path(config$outputs$figures, "incidence_rates.jpg"), 
#        scale=0.7, width=25, height=15, units="cm")

```

# Yearly incidence rates in UKB as sanity check

```{r}

data_wholepop <- wholepop %>% transmute(
  TEU_BrCa_status = as.numeric(!is.na(BrCaAge_Any) | TEU_BrCa_status),
  TEU_BrCa_age = coalesce(BrCaAge_Any, TEU_BrCa_age), # Take CaR recorded age if exists
  TEU_BaC_AgeAtRec = TEU_BaC_AgeAtRec
) %>%
  filter(TEU_BrCa_age >= 40) # Making this explicit

data_analysis <- analysis_pop %>% select(
  TEU_BrCa_status,
  TEU_BrCa_age,
  TEU_BaC_AgeAtRec
)

# Yearly incidence
cases_wholepop <- c()
cases_analysispop <- c()
people_wholepop <- c()
people_analysispop <- c()

for(age in seq(40, 84, by=1)){
  
 cases_wholepop <- c(cases_wholepop, nrow(data_wholepop %>% filter(TEU_BrCa_status == 1 & round(TEU_BrCa_age) == age)))
 cases_analysispop <- c(cases_analysispop, nrow(data_analysis %>% filter(TEU_BrCa_status == 1 & round(TEU_BrCa_age) == age)))
 
 people_wholepop <- c(people_wholepop, nrow(data_wholepop %>% filter(TEU_BrCa_age >= age)))
 people_analysispop <- c(people_analysispop, nrow(data_analysis %>% filter(TEU_BaC_AgeAtRec <= age & TEU_BrCa_age >= age)))
}

rates <- data.frame(age = seq(40, 84, by=1),
                    cases_wholepop = cases_wholepop,
                    cases_analysispop = cases_analysispop,
                    people_wholepop = people_wholepop,
                    people_analysispop = people_analysispop) %>%
  mutate(rates_wholepop = cases_wholepop/(people_wholepop/100000),
         rates_analysispop = cases_analysispop/(people_analysispop/100000),
         rates_ons = rep(c(125.3, 230.4, 281.0, 276.0, 354.8, 415.9, 354.9, 400.6, 432.1), each=5)) %>%
  pivot_longer(cols=c(rates_analysispop, rates_wholepop, rates_ons)) %>%
  mutate(Cohort = factor(name, levels=c("rates_ons", "rates_wholepop", "rates_analysispop"),
                       c("ONS 2013", "Whole UK Biobank", "Analysis population")))

ggplot(data=rates, aes(x=age, y=value, group=Cohort, color=Cohort)) +
  geom_line() +
  geom_point() +
  ggtitle("Age specific breast cancer rates in women in the UK Biobank,\noverall and within analysis cohort, compared to\nOffice for National Statistics 2013 Cancer Registry data") +
  ylab("Incidence rate per 100,000 person years at risk") +
  xlab("Age at cancer diagnosis, years") 

```


